Curve counting in genus one: elliptic singularities {\it \&} relative geometry

### 2019-06-28 EllipticSingularitiesSplitting.tex

Our analysis leads naturally to the study of a ``main component’’ double ramification cycle for target manifolds in genus one. The virtual geometry of this was studied in recent work of Janda, Pandharipande, Pixton, and Zvonkine~\cite{DRCBundle}. We do not know how to calculate a closed formula for this main component contribution, but we repurpose E.~Katz's topological recursion relations \cite{EKatz} to complete the algorithm. We note that even in genus one, an understanding of the main component contributions of the double ramification cycle for target manifolds would be interesting.

@@

\subsection*{Acknowledgements} We have benefited from fruitful conversations with friends and colleagues, including Francesca Carocci, Davesh Maulik, Jonathan Wise, and \tred{[other people]}. Important aspects of this work were completed while L.B. and N.N. were visiting MIT in Spring 2018 and the University of Cambridge in Winter 2019, and we thank these institutions for their hospitality. \tred{L.B. would like to thank Imperial College London, the LSGNT, the Royal Society, 3CinG, the MSRI program ''Enumerative Geometry beyond Numbers``, and MPIM-Bonn for support.}

@@

The above theorem guarantees that the singularities of the space of totally factorised maps to expansions is logarithmically smooth. In fact, one can say more in this case.\marginpar{I don't understand the point of this Cor. Can we make it into a short remark?}

@@

Equipped with the strata logarithmic structures, the morphism
\[
\widetilde \Dcal\to \Ecal
\]
is a logarithmic morphism. The generic fibre of this morphism is toric variety.  \medskip

@@

This procedure induces a logarithmic modification $\widetilde\Ecal \to \Ecal$ which admits a modular description as a fibre product of genus zero centrally aligned maps coming from the vertices $\sqC_1,\ldots,\sqC_r$, and genus one aligned rubber maps satisfying fibre-direction factorisation from the vertex $\sqC_0$). We have thus produced a factorisation
\begin{equation*}\widetilde\Dcal\to\widetilde \Ecal\to\Ecal,
\end{equation*}
where the first map is flat and the second map is a logarithmic modification. $\widetilde\Dcal$ is a compactification of a torus bundle over $\widetilde\Ecal$ whose strata can be read off from the tropicalization; in particular, we have the information of how (polynomials in) the classes of boundary strata push forward to $\widetilde\Ecal$. In fact, the generic fibre of $\widetilde\Dcal \to \widetilde\Ecal$ can be read off geometrically.

\begin{lemma}
 Generically on $\Ecal$, $\widetilde \Dcal\to\widetilde\Ecal$ is isomorphic to $\mathcal P=\PP_{\widetilde\Ecal}(\bigoplus_{v\in \{V_{0,\delta}\}\cup V_{>0,\delta}}\OO(\delta_v))$.
\end{lemma}

\begin{proof}
 All of the line bundles $\OO(\delta_v)$ share a common factor $\mathbb T_E$, the universal tangent line to the elliptic curve. By removing this factor, we see that when the dual graph is fixed, $\OO(\delta_v)$ is isomorphic to the tangent line to the rational tail $R_v$ at the gluing node $q_v$. We make the caveat here that(with the caveat that all those corresponding to $V_{0,\delta}$ are canonically identified). This is the open stratum of the moduli space of attaching data \cite[\S\S 2.2-2.3]{SMY2}. On the fibre boundary, the circle of radius $\delta$ will pass through a subset of $\{V_{0,\delta}\}\cup V_{>0,\delta}$, while cutting the complementary subset along the edge between the core and $v$; correspondingly, there is a destabilisation of the curve, and an elliptic singularity containing strictly semistable branches. Tropically, pushing some of the vertices off the circle creates new finite-length edges with coordinates $\ell_v\in\RR_{\geq_0}$ (see Figure \ref{fig:off_we_go}); not all of them can be strictly positive, because the circle needs to pass through at least one vertex (of positive horizontal degree). This is patently the fan of a projective space.
\end{proof}

@@

\subsection{Recursion for general $(X,Y)$}\label{section recursion for general pair} We now let $(X,Y)$ be a general smooth very ample pair. In \S \ref{subsection virtual} we studied the moduli space $\VZ_{1,\alpha}(X|Y,\beta)$ of centrally aligned maps to $(X,Y)$ and equipped it with a virtual fundamental class by diagonal pull-back from $\VZ_{1,\alpha}(\PP^N|H,d)$.

>>>>

\tred{We rephrase Gathmann's idea of increasing the tangency requirement within the framework of logarithmic stable maps (where maximal tangency is always assumed) by introducing \emph{fictitious markings}, and identifying the space of maps with higher contact order with a boundary stratum inside that of lower contact order and more (fictitious) markings.}

Our analysis leads naturally to \tred{the} study of a ``main component’’ double ramification cycle for target manifolds in genus one. The virtual geometry of this was studied in recent work of Janda, Pandharipande, Pixton, and Zvonkine~\cite{DRCBundle}. We do not know how to calculate a \tred{closed formula} for this main component contribution, but we repurpose E.~Katz's topological recursion relations \cite{EKatz} to complete the algorithm. We note that even in genus one, an understanding of the main component contributions of the double ramification cycle for target manifolds would be interesting.

@@

\subsection*{Acknowledgements} We have benefited from fruitful conversations with friends and colleagues, including Francesca Carocci, Cristina Manolache, Davesh Maulik, Jonathan Wise, and [other people]. Important aspects of this work were completed while L.B. and N.N. were visiting MIT in Spring 2018 and the University of Cambridge in Winter 2019, and we thank these institutions for their hospitality. L.B. would like to thank Imperial College London, the LSGNT, the Royal Society, 3CinG, the MSRI program ''Enumerative Geometry beyond Numbers``, and MPIM-Bonn for support.

@@

The above theorem guarantees that the singularities of the space of totally factorised maps to expansions is logarithmically smooth. In fact, one can say more in this case.

@@

Equipped with the strata logarithmic structures, the morphism
\[
\widetilde \Dcal\to \Ecal
\]

is a logarithmic morphism. The generic fiber of this morphism is a toric variety.  \medskip

@@

This procedure induces a logarithmic modification $\widetilde\Ecal \to \Ecal$ which admits a modular description as a fibre product of genus zero centrally aligned maps coming from the vertices $\sqC_1,\ldots,\sqC_r$, and genus one aligned rubber maps satisfying fibre-direction factorisation from the vertex $\sqC_0$.

By construction, we have a map $\widetilde\Dcal \to \widetilde\Ecal$ which is a compactification of a torus bundle over $\widetilde\Ecal$. The strata of $\widetilde\Dcal$ can be read off from its tropicalisation, and in particular we have the information of how (polynomials in) the classes of boundary strata push forward to $\widetilde\Ecal$.

We have produced a factorization
\begin{equation*}\widetilde\Dcal\to\widetilde \Ecal\to\Ecal,
\end{equation*}
where the first map is flat and the second map is a logarithmic modification. $\widetilde\Dcal$ is a compactification of a torus bundle over $\widetilde\Ecal$ whose strata can be read off from the tropicalization; in particular, we have the information of how (polynomials in) the classes of boundary strata push forward to $\widetilde\Ecal$. In fact, the generic fibre of $\widetilde\Dcal \to \widetilde\Ecal$ can be read off geometrically.

\begin{lemma}\label{lem:generic_proj_bundle}
 Generically on $\Ecal$, $\widetilde \Dcal\to\widetilde\Ecal$ is given by an explicit projective bundle.
\end{lemma}

\begin{proof}
Let us assume that the curve does not degenerate, so that the dual graph is as prescribed by $\Delta$, in particular there are components $\sqC_1,\ldots,\sqC_r$ generically lying on the circle of radius $\delta$; the moduli space of attachments for the elliptic singularity is the complement of the coordinate hyperplanes in $\PP(\bigoplus_{i=1}^rT_{q_i}C_i)$. On the boundary, the circle is allowed to miss some of the $\sqC_i$ - but not all of them! -, cutting the edge between them and the core (correspondingly, there is a destabilisation of the curve, and an elliptic singularity containing strictly semistable branches). Tropically, pushing some of the vertices off the circle creates new finite-length edges with coordinates $\ell_i\in\RR_{\geq_0}$ (see Figure \ref{fig:off_we_go}). Since not all of them can be strictly positive, this represents the fan of a projective space.

A slight modification of this argument is required to account for the fact that all vertices that lie on the circle and are mapped to $0\in\mathbb R_{\geq0}$ by the tropical map are generically aligned on $\Ecal$: indeed, by Lemma \ref{lemma type C0 combinatorial types}, they all have the same contact order $m$, therefore their $T_{q_i}C_i^{\otimes m}$ can be identified via the target tangent line on the fiber product \eqref{IIIa fibre product}, and, after saturation, the $T_{q_i}C_i$ themselves are canonically isomorphic on $\Ecal$. Thus, only one factor should be taken to represent all of them in the projective bundle above.
\end{proof}

@@

To establish a more direct connection with the geometry of the fibre product \eqref{IIIa fibre product}, recall that $\widetilde\Ecal$ has been built as an explicit modification of a finite cover of \eqref{IIIa fibre product}. For every $v\in\{\sqC_1,\ldots,\sqC_r\}$ - with the convention established at the end of Lemma \ref{lem:generic_proj_bundle} -, on $\widetilde\Ecal$ it makes sense to consider the line bundle $\OO(\delta_v)$ associated to a shortest path from the core to the circle in the $v$-direction. Construct $\mathcal P=\PP_{\widetilde\Ecal}(\bigoplus_{v\in \{\sqC_1,\ldots,\sqC_r\}}\OO(\delta_v))$; since $\sqC_1,\ldots,\sqC_r$ are always ordered on $\widetilde \Dcal$ (they may lie on the circle, be the first component off the circle, or degenerate and lie in the strict interior), there is a morphism $\widetilde \Dcal\to\mathcal P$ lifting $\widetilde \Dcal\to\widetilde\Ecal$. Consider the map of vector bundles on $\mathcal P$ (analogous to \eqref{eq:factorisation}):
\[ \OO_{\mathcal P}(-1)\to p^*(\bigoplus_{v\in \{\sqC_1,\ldots,\sqC_r\}}\OO(\delta_v))\xrightarrow{\sum\operatorname{d}f}\ev_{C_0}^*T_H.\]

The composite, regarded as a section $s$ of $G=\OO_{\mathcal P}(1)\otimes \operatorname{ev}_{C_0}^*(T_H)$, vanishes precisely where $f$ factors through the Smyth's singularity determined by the point in the fiber of $p$. Generically on $\widetilde\Ecal$, this is exactly what we are after - compare with Lemma \ref{lem:generic_proj_bundle}. On the boundary, though, it may happen that a simultaneous degeneration in the base and the fiber makes this factorisation trivial, in the sense that $\bar f$ would be constant on all the branches of the elliptic singularity.

\begin{lemma}
 For every $\Lambda_B\subseteq \{\sqC_1,\ldots,\sqC_r\}$ and $\Lambda_F=\{\sqC_1,\ldots,\sqC_r\}\setminus \Lambda_B$, $s$ vanishes along the loci where the factors corresponding to $\Lambda_B$ in $\widetilde \Ecal$ degenerate so that $f_i\colon C_i\to\PP^N$ is constant in a neighbourhood of $q_i$, and the fiber coordinates $(\lambda_j)_{j\in\Lambda_F}$ along $p\colon\mathcal P\to\widetilde \Ecal$ are set to be $0$.
\end{lemma}
Call these loci $\Xi_i,i\in I$ (compare with \cite[\S 3.2]{VZ} for an analogous phenomenon in the non-relative case); clearly, they may be in excess with respect to the expected codimension of $V(s)$, that is $N-1$. These loci are logarithmic substrata; therefore, there exists a logarithmic blow-up that principalizes them. We claim that $\widetilde\Dcal$ is (possibly a refinement of) such a blow-up. 
\begin{lemma}
 The pullback of $\Xi_i$ is principal in $\widetilde\Dcal$.
\end{lemma}
\begin{proof}
 Let us analyse the ideal of $\Xi_i$: it is generated by
 \begin{enumerate}[leftmargin=.5 cm]
  \item equations on $\widetilde\Ecal$ cutting the locus where $C_i$ degenerates so that $f_i$ is constant on a neighbourhood of $q_v$, $i\in \Lambda_B$: these correspond to the tropical functions $\delta_i-\operatorname{dist}(q_i,\circ)$;
  \item equations on the fiber of $p$ for the linear subspace corresponding to $j\in \Lambda_F$: tropically, these correspond to the $\ell_j$ discussed above.
 \end{enumerate}
By construction, on $\widetilde \Dcal$ it is always possible to tell which of these tropical lengths is the shortest (the circle we have drawn on $\Xi$ is degenerate - it passes through no vertex of positive horizontal degree -, therefore we need to decide for a strictly larger circle in order to lift to $\widetilde D$), which corresponds to principalising the ideal (see Figure \ref{fig:principalisation}).
\end{proof}

\begin{figure}
\includegraphics[scale=.5]{principalisation}
\caption{On the left, a generic point of $\mathcal P$. On the right, a point of $\Xi$. The dotted circle is the one we draw on $\mathcal P$. The continuous circle represents one of the possible choices of $\delta$ enlarging the dotted one (a stratum of $\widetilde\Dcal$).}\label{fig:principalisation}
\end{figure}

Finally, on $\widetilde\Dcal$ (or, in fact, on an intermediate blow-up $\widetilde{\mathcal P}\to\mathcal P$) there are exceptional divisors $\widetilde \Xi_i$ corresponding to $\Xi_i$ in $\mathcal P$. Since the $\Xi_i$ exhaust all the excess-dimensional components of the vanishing locus of $s\in\Gamma(\mathcal P,G)$, the latter induces a section $\tilde s\in\Gamma(\widetilde\Dcal,G(-\sum_i\widetilde\Xi_i))$ by pullback and twisting, whose vanishing locus is dimensionally transverse to the boundary, and whose open part coincides with the generic point of $\Dcal$, therefore $V(\tilde s)=\Dcal$.

Expressing the pushforward of the Chern classes of $\OO_F(1)$ and $\OO(\widetilde\Xi_i)$ in terms of tautological classes on the fiber product underlying $\Ecal$, we get:
\begin{cor}
 After pushforward to $\Ecal$, the contribution of $\Dcal$ to the invariants can be computed in terms of tautological integrals on moduli spaces of maps with lower numerics.
\end{cor}

\subsection{Recursion for general $(X,Y)$}\label{section recursion for general pair} We now let $(X,Y)$ be a general smooth very ample pair. In \S \ref{} we studied the moduli space $\VZ_{1,\alpha}(X|Y,\beta)$ of centrally aligned maps to $(X,Y)$ and equipped it with a virtual fundamental class by diagonal pull-back from $\VZ_{1,\alpha}(\PP^N|H,d)$.
